Wait-free Queues with Polygarithmic Step Complexity

### 2022-05-13 article/psuedocode.tex
add proof for some of the first lemmas

\Function{boolean}{Refresh()}{}
\State \tt{s= size} \label{readSize}
%\If{\tt{n.blocks[h]!=null?}} \tt{h+=1}\EndIf
\State \tt{<new, np\sub{left}, np\sub{right}>= \Call{CreateBlock}{h}} \cmt{\tt{np\sub{left}, np\sub{right}} are the values read from the children's \tt{num\sub{propagated}} feild.}
\If{\tt{new.num==0}} \Return{\tt{true}} \cmt{The block contains nothing.}
\ElsIf{\tt{blocks.tryAppend(new, s)}}
\ForEach{\tt{dir} {\keywordfont{in}} \tt{\{left, right\}}} \label{okcas}
\State \tt{\Call{CAS}{dir.super[np\sub{dir}], null, h+1}} \cmt{Write would work too.}
\State \tt{\Call{CAS}{dir.num\sub{propagated}, np\sub{dir}, np\sub{dir}+1}}
\EndFor
\State \tt{\Call{CAS}{size, h, h+1}}
\State \Return{\tt{true}}
\Else
\State \tt{\Call{CAS}{size, h, h+1}} \cmt{Even if another process wins, help to increase the size. The winner might have fallen sleep before increasing \tt{size}.}
\State \Return{ \tt{false}}
\EndIf
\EndFunction{Refresh}

@@

@@

\begin{lemma}[headProgress] \label{lem::sizeProgress}
\nf{n.established} is non-decreasing over time.
\end{lemma}
%\begin{proof}
%  Simply because \nf{n.head} is only incremented.
%\end{proof}

\begin{lemma}[headPosition] \label{lem::headPosition} The value read in Line \ref{prevLine}(\nf{h=n.established}) might be 1 bit behind the first empty block in the node.
\end{lemma}
\begin{proof}
 Because at the end of every \nf{Refresh()} with block size greater than 0 (Lines 53,56) \nf{n.head} is incremented. Maybe some process goes to sleep before incrementing the head, but after sleeping if \nf{h} does not increase then \nf{CAS} in Line 52 is going to be failed and nothing is going to be appended to \nf{n.blocks}.
\end{proof}

@@

\begin{lemma}[establishedOrder]\label{lem::establishedOrder}
  If  time $t<$ time $t^\prime$, then \nf{ops(EST\sub{n, t})}$\subseteq$ \nf{ops(EST\sub{n, t$^\prime$})}.
\end{lemma}
\begin{proof}
Because blocks are only appended(not modified) with CAS to \nf{n.blocks[n.head]} and \nf{n.head} is non-decreasing.
\end{proof}

\begin{lemma}[createBlock] \label{lem::createBlock}
  If \nf{b} is the block returned by \nf{n.CreateBlock(h, x)} invoked at time $t$, then \nf{ops(EST\sub{n.left, t})} $\cup$ \nf{ops(EST\sub{n.right, t})} - \nf{ops(EST\sub{n, t})} $\subseteq$ \nf{ops(b)}.
\end{lemma}
\begin{proof}
 We prove the claim for the left child. Blocks in \nf{n.left.blocks[n.blocks[i-1].end\textsubscript{left}+1..n.blocks[i].end\textsubscript{left}]} are all the new established operations at time $t$ by definition of Subblock. Line 70 is after $t$ and since the head is only increasing (Lemma~\ref{lem::headProgress}) the lemma holds. See Figure~\ref{fig::createBlock}. The right child is the same.
\end{proof}

\begin{lemma}[trueRefresh] \label{lem::trueRefresh}
  Let $t_i$ be the time \nf{n.Refresh()} is invoked and $t_t$ be the time it is terminated. Suppose \nf{n.Refresh()}'s \nf{TryAppend(new, s)} returns \nf{true}, then \nf{ops(EST\sub{n.left, t\sub{i}})} $\cup$ \nf{ops(EST\sub{n.right, t\sub{i}})} $\in$ \nf{ops(EST\sub{n, t\sub{t}})}.
\end{lemma}
\begin{proof}
By Lemma~\ref{lem::createBlock} \nf{new} contains \nf{n}'s childrens' established blocks before Line 43 which is  appended to \nf{n.blocks} by \nf{CAS} in Line 48.
\end{proof}

\begin{lemma}[falseRefresh] \label{lem:falseRefresh}
    If instance \nf{r} of \nf{n.Refresh()} reads value \nf{s} on line \ref{readSize} and then returns \nf{false}, then there is another instance $r^\prime$ of \nf{n.Refresh()} that has performed a successful \nf{TryAppend(new, s)}. A \nf{TryAppend()} is successful if its \nf{CAS} is successful.
\end{lemma}
\begin{proof}
  If there is no other concurrent successful \nf{Refresh(n)} then \nf{Refresh(n)} would succeed in Line 48. So there is another \nf{Refresh(n)}, that has to \nf{CAS}ed successfully its block in \nf{n.blocks[h]} after Line 43 of \nf{Refresh(n)}. Otherwise the other \nf{Refresh(n)} should have read \nf{h$^\prime$>h} instead of \nf{h} for \nf{n.head}(Line 52).
\end{proof}

\begin{lemma}[Double Refresh] \label{doubleRefresh}
  Consider two consecutive failed instances $r_1$,$r_2$ of \nf{n.Refresh()} by some process. Let $t_1$ be the time $r_1$ is invoked and $t_2$ be the time $r_2$ terminated. After $r_2$'s \nf{TryAppend} we have \nf{ops(EST\sub{n.left, t\sub{1}})} $\cup$ \nf{ops(EST\sub{n.right, t\sub{1}})} $\subseteq$ \nf{ops(EST\sub{n, t\sub{2}})}.
\end{lemma}
\begin{proof}

If Line 35 (first Refresh which we call $R_1$) returns \nf{true}, the claim is held by Lemma~\ref{lem::trueRefresh}. If not, then there is another successful instance of \nf{Refresh()} $R_1^\prime$ by Lemma~\ref{lem:falseRefresh}. $R_1^\prime$ may or may not have propagated some subblocks of \nf{new} to \nf{n}. It is obvious that the \nf{new} constructed by the second Refresh in Line 36 contains the blocks in \nf{new} by $R_1$ which $R_1^\prime$ did not contain, since \nf{n.head} is only increasing (Maybe Lemma~\ref{lem::oldnewOrder}). If $R_2$ succeeds by Lemma~\ref{lem::trueRefresh} the claim holds. If not, it is deduced that \nf{n.blocks[h]} was not \nf{null} before $R_2$'s \nf{CAS}. Furthermore, \nf{n.blocks[h]} was \nf{null} before reading \nf{h} by $R_1$. So there is a successful \nf{Refresh()} after the read of \nf{h} in $R_1$ and before the \nf{CAS} of $R_2$. This \nf{Refresh()} contains all the new established operations before Line 35 (Maybe Lemma~\ref{lem::oldnewOrder}) and by Lemma~\ref{lem::trueRefresh} our claim holds. See Figure~\ref{}.
\end{proof}

@@

\begin{corollary}[Propagate Step] \label{doublyRefresh}
All operations in \nf{n}'s children's established blocks before line \ref{firstRefresh} are guaranteed to be in \nf{n}'s blocks after line~\ref{secondRefresh}.
\end{corollary}
\begin{proof}
Suppose block \nf{b} with index \nf{i} is in the the left child of \nf{n} before the line 35. By Lemma~\ref{head} it follows that \nf{n.left.head} is greater than \nf{i}. \nf{Refresh()} calls \nf{CreateBlock()} and creates a block from blocks between \nf{n.blocks[n.head].end\textsubscript{left}} and \nf{n.left.head} in the left child, which contains \nf{b} as well. First it tries to append it in \nf{n.blocks.head} and if it was succsuful it continues recursively. If not it tries again, and if the second call of \nf{Refresh()} in Line 36 fails. It means there is another \nf{Refresh} which has reah its \nf{i} after the Line 35, so it contains \nf{b} as well.
\end{proof}
\nf{CreateBlock()} reads blocks in the children that do not exist in the parent and aggregates them into one block. If a \nf{Refresh()} procedure returns true it means it has appended the block created by \nf{CreateBlock()} into the parent node's sequence. So suppose two \nf{Refresh}es fail. Since the first \nf{Refresh()} was not successful, it means another CAS operation by a \nf{Refresh}, concurrent to the first \nf{Refresh()}, was successful before the second \nf{Refresh()}. So it means the second failed \nf{Refresh} is concurrent with a successful \nf{Refresh()} that assuredly has read block before the mentioned line \nf{35}. After all it means if any of the \nf{Refresh()} attempts were successful the claim is true, and also if both fail the mentioned claim still holds.

\begin{lemma}[No Duplicates]\label{append}
If \nf{op} is appended to \nf{n.blocks[i]} then after that there is no \nf{j>i} such that \nf{op}$\in$\nf{ops(n.blocks[j])}.
\end{lemma}
\begin{proof}
\nf{Append(op)} adds \nf{op} to \nf{l\textsubscript{p}.blocks}(Line~\ref{addOP}) and \nf{Propagate()} recursively propagates \nf{op} up to the root.
By lemma \ref{doublyRefresh} we know that operation \nf{op} propagates from child to parent at each level.
\end{proof}
\begin{corollary}
  After \nf{Append(blk)} finishes \nf{ops(blk)}$\subseteq$\nf{ops(root.blocks[x])} for some \nf{x} and only one \nf{x}.
\end{corollary}

@@

\framebox[1.1\width]{TODO} Fallback safety lemmas.

>>>>

\Function{boolean}{Refresh()}{}
\State \tt{s= size} \label{readSize}
%\If{\tt{n.blocks[h]!=null?}} \tt{h+=1}\EndIf
\State \tt{<new, np\sub{left}, np\sub{right}>= \Call{CreateBlock}{h}} \cmt{\tt{np\sub{left}, np\sub{right}} are the values read from the children's \tt{num\sub{propagated}} feild.}
\If{\tt{new.num==0}} \Return{\tt{true}} \cmt{The block contains nothing.}
\ElsIf{\tt{blocks.tryAppend(new, s)}}
\ForEach{\tt{dir} {\keywordfont{in}} \tt{\{left, right\}}} \label{okcas}
\State \tt{\Call{CAS}{dir.super[np\sub{dir}], null, h+1}} \cmt{Write would work too.}
\State \tt{\Call{CAS}{dir.num\sub{propagated}, np\sub{dir}, np\sub{dir}+1}}
\EndFor
\State \tt{\Call{CAS}{size, h, h+1}} \label{incrementHead1}
\State \Return{\tt{true}}
\Else
\State \tt{\Call{CAS}{size, h, h+1}} \cmt{Even if another process wins, help to increase the size. The winner might have fallen sleep before increasing \tt{size}.} \label{incrementHead2}
\State \Return{ \tt{false}}
\EndIf
\EndFunction{Refresh}

@@

\framebox[1.1\width]{TODO} Fallback safety lemmas. Some parts are obsolete.

\framebox[1.1\width]{Questions} When I write the lemmas since every claim in my mind is correct maybe I miss some fact that need proof or maybe I refer to some lemmas that are generally correct but not needed for the linearizability proof. Is lemma 7 necessary? Is lemma 13 induced trivially from lemma 8?

@@

\begin{lemma}[headProgress] \label{lem::sizeProgress}
\nf{n.established} is non-decreasing over time.
\end{lemma}
\begin{proof}
  Induced trivially from the pseudocode since \nf{n.established} is only incremented.
\end{proof}

\begin{lemma}[headPosition] \label{lem::headPosition} If the value read in Line \ref{prevLine}(\nf{h=n.established}) is $h$ then, \nf{n.blocks[i]=null} for \nf{i>h} and \nf{n.blocks[i]$\neq$null} for \nf{i<h}.
\end{lemma}
\begin{proof}
At the end of every \nf{n.Refresh()} with a block  with size greater than 0 returned by \nf{CreateBlock()} \nf{n.head} is incremented (Lines \ref{incrementHead1}, \ref{incrementHead2}). If a process went to sleep before incrementing the \nf{established} (line \ref{incrementHead1}), nothing can be appended to \nf{n.blocks} until another process increments \nf{n.establsihed} (Line \ref{incrementHead2}).
\end{proof}

@@

\begin{lemma}[establishedOrder]\label{lem::establishedOrder}
  If  time $t<$ time $t^\prime$, then \nf{ops(EST\sub{n, t})}$\subseteq$ \nf{ops(EST\sub{n, t$^\prime$})}.
\end{lemma}
\begin{proof}
Blocks are only appended(not modified) with CAS to \nf{n.blocks[n.head]} and \nf{n.head} is non-decreasing, so the set of operations in established blocks of a node grows.
\end{proof}

\begin{lemma}[createBlock] \label{lem::createBlock}
  If \nf{b} is the block returned by \nf{n.CreateBlock(h, x)} invoked at time $t$, then \nf{ops(EST\sub{n.left, t})} $\cup$ \nf{ops(EST\sub{n.right, t})} - \nf{ops(EST\sub{n, t})} $\subseteq$ \nf{ops(b)}.
\end{lemma}
\begin{proof}
 WLOG we prove the claim for the left child. Blocks in \nf{n.left.blocks[n.blocks[i-1].end\textsubscript{left}+1..n.blocks[i].end\textsubscript{left}]} are all the new established operations at time $t$ by definition of Subblock. Since the head is only increasing (Lemma~\ref{lem::sizeProgress}) the  claim holds after $t$. See Figure~\ref{fig::createBlock}. The right child is the same.
\end{proof}

\begin{lemma}[trueRefresh] \label{lem::trueRefresh}
  Let $t_i$ be the time \nf{n.Refresh()} is invoked and $t_t$ be the time it is terminated. Suppose \nf{n.Refresh()}'s \nf{TryAppend(new, s)} returns \nf{true}, then \nf{ops(EST\sub{n.left, t\sub{i}})} $\cup$ \nf{ops(EST\sub{n.right, t\sub{i}})} $\in$ \nf{ops(EST\sub{n, t\sub{t}})}.
\end{lemma}
\begin{proof}
By Lemma~\ref{lem::createBlock} \nf{new} contains \nf{n}'s childrens' established blocks after Line \ref{readSize} which is  appended to \nf{n.blocks[]} by \nf{TryAppend} in Line 313.
\end{proof}

\begin{lemma}[falseRefresh] \label{lem:falseRefresh}
    If instance \nf{r} of \nf{n.Refresh()} reads value \nf{s} on line \ref{readSize} and then returns \nf{false}, then there is another instance $r^\prime$ of \nf{n.Refresh()} that has performed a successful \nf{TryAppend(new, s)}. A \nf{TryAppend()} is successful if its \nf{CAS} is successful.
\end{lemma}
\begin{proof}
  If there is no other concurrent successful \nf{n.Refresh()} then \nf{n.Refresh()} would succeed in Line 313. So there is another \nf{n.Refresh()}, that has appended its block in \nf{n.blocks[h]} after Line 310 of the first \nf{n.Refresh()}. 
%Otherwise the other \nf{Refresh(n)} should have read \nf{h$^\prime$>h} instead of \nf{h} for \nf{n.head}(Line 52).
\end{proof}

\begin{lemma}[Double Refresh] \label{doubleRefresh}
  Consider two consecutive failed instances $r_1$,$r_2$ of \nf{n.Refresh()} by some process. Let $t_1$ be the time $r_1$ is invoked and $t_2$ be the time $r_2$ terminated. After $r_2$'s \nf{TryAppend} we have \nf{ops(EST\sub{n.left, t\sub{1}})} $\cup$ \nf{ops(EST\sub{n.right, t\sub{1}})} $\subseteq$ \nf{ops(EST\sub{n, t\sub{2}})}.
\end{lemma}
%\begin{proof}
%  
%
%If Line 35 (first Refresh which we call $R_1$) returns \nf{true}, the claim is held by Lemma~\ref{lem::trueRefresh}. If not, then there is another successful instance of \nf{Refresh()} $R_1^\prime$ by Lemma~\ref{lem:falseRefresh}. $R_1^\prime$ may or may not have propagated some subblocks of \nf{new} to \nf{n}. It is obvious that the \nf{new} constructed by the second Refresh in Line 36 contains the blocks in \nf{new} by $R_1$ which $R_1^\prime$ did not contain, since \nf{n.head} is only increasing (Maybe Lemma~\ref{lem::oldnewOrder}). If $R_2$ succeeds by Lemma~\ref{lem::trueRefresh} the claim holds. If not, it is deduced that \nf{n.blocks[h]} was not \nf{null} before $R_2$'s \nf{CAS}. Furthermore, \nf{n.blocks[h]} was \nf{null} before reading \nf{h} by $R_1$. So there is a successful \nf{Refresh()} after the read of \nf{h} in $R_1$ and before the \nf{CAS} of $R_2$. This \nf{Refresh()} contains all the new established operations before Line 35 (Maybe Lemma~\ref{lem::oldnewOrder}) and by Lemma~\ref{lem::trueRefresh} our claim holds. See Figure~\ref{}.
%\end{proof}

@@

\begin{corollary}[Propagate Step] \label{doublyRefresh}
All operations in \nf{n}'s children's established blocks before line \ref{firstRefresh} are guaranteed to be in \nf{n}'s blocks after line~\ref{secondRefresh}.
\end{corollary}
%\begin{proof}
%Suppose block \nf{b} with index \nf{i} is in the the left child of \nf{n} before the line 35. By Lemma~\ref{head} it follows that \nf{n.left.head} is greater than \nf{i}. \nf{Refresh()} calls \nf{CreateBlock()} and creates a block from blocks between \nf{n.blocks[n.head].end\textsubscript{left}} and \nf{n.left.head} in the left child, which contains \nf{b} as well. First it tries to append it in \nf{n.blocks.head} and if it was succsuful it continues recursively. If not it tries again, and if the second call of \nf{Refresh()} in Line 36 fails. It means there is another \nf{Refresh} which has reah its \nf{i} after the Line 35, so it contains \nf{b} as well.
%\end{proof}

\nf{CreateBlock()} reads blocks in the children that do not exist in the parent and aggregates them into one block. If a \nf{Refresh()} procedure returns true it means it has appended the block created by \nf{CreateBlock()} into the parent node's sequence. So suppose two \nf{Refresh}es fail. Since the first \nf{Refresh()} was not successful, it means another CAS operation by a \nf{Refresh}, concurrent to the first \nf{Refresh()}, was successful before the second \nf{Refresh()}. So it means the second failed \nf{Refresh} is concurrent with a successful \nf{Refresh()} that assuredly has read block before the mentioned line \nf{35}. After all it means if any of the \nf{Refresh()} attempts were successful the claim is true, and also if both fail the mentioned claim still holds.

\begin{lemma}[No Duplicates]\label{append}
If \nf{op} is appended to \nf{n.blocks[i]} then after that there is no \nf{j>i} such that \nf{op}$\in$\nf{ops(n.blocks[j])}.
\end{lemma}
%\begin{proof}
%\nf{Append(op)} adds \nf{op} to \nf{l\textsubscript{p}.blocks}(Line~\ref{addOP}) and \nf{Propagate()} recursively propagates \nf{op} up to the root.
%By lemma \ref{doublyRefresh} we know that operation \nf{op} propagates from child to parent at each level.
%\end{proof}

\begin{corollary}
  After \nf{Append(blk)} finishes \nf{ops(blk)}$\subseteq$\nf{ops(root.blocks[x])} for some \nf{x} and only one \nf{x}.
\end{corollary}

@@
